{\bf Affine.m} -- {\it Mathematica} package for computations in representation theory of finite-dimensional and affine Lie algebras

### 2011-07-13 doc/paper.tex
Notes from paper added

\begin{abstract}
%% Text of abstract
We present {\bf Affine.m} -- program for computations in representation theory of finite-dimensional and affine Lie algebras. It can be used to study irreducible, Verma and parabolic Verma modules. Algorithms are based upon the properties of weights and Weyl symmetry. Weight multiplicity computation, branching of representation to representations of subalgebra and tensor product decomposition are the most important problems that we treat. These problems has numerous applications in physics.

The program is implemented in popular  computer algebra system {\it Mathematica} and works with finite-dimensional and affine Lie algebras.

@@

{\it Lie algebra} $\gf$ is a vector space with bilinear operation $[\cdot,\cdot]:\gf\otimes\gf\to \gf$, which is called {\it commutator}. If we choose the basis $X_{i}$ in $\gf$ we can specify commutation relations by the {\it structure constants} $C_{ijk}$:
\begin{equation}
  \label{eq:1}
  [X^{i},X^{j}]=\sum_{k} C^{ij}_{k} X^{k}
\end{equation}
Lie algebra is simple if it contains no non-trivial ideals. Semisimple Lie algebra is a direct sum of simple Lie algebras.

Maximal commutative subalgebra (Cartan subalgebra) of $\gf$ is denoted by $\hfg$.
We denote the elements of basis of $\hfg$ by $H^{i}$.

@@

Although the full set of roots $\Delta$ is infinite for affine Lie algebra $\hat\gf$ the set of simple roots $\Phi$ is finite and its elements are denoted by $\alpha_{0},\dots \alpha_{r}$ where $r=\mathrm{rank}(\gf)$. The roots $\alpha_1,\dots, \alpha_r$ are the roots of the underlying finite-dimensional Lie algebra $\go$. The root $\alpha_0=\delta-\theta$ is the difference of {\it imaginary root} $\delta$ and $\theta$ -- highest root of the algebra $\go$.

@@

The Weyl group $W_{\gf}$ is generated by reflections $\{s_{i}:\hfg^{*}\to\hfg^{*}\}$ corresponding to simple roots $\{\alpha_{i}\}$:
\begin{equation}
  \label{eq:8}
  s_{i}\cdot\lambda=\lambda-\frac{2(\alpha_{i},\lambda)}{(\alpha_{i},\alpha_{i})}\alpha_{i}
\end{equation}
Note that root multiplicity $\mathrm{mult}(\alpha)$ for affine Lie algebra can be greater than one.

Root system and characters of representation are invariant with respect to the action of Weyl group. Root system can be reconstructed from the set of simple roots with the action of Weyl group.

@@

Consider an action of element $s_{0}s_{i}s_{0}$ of Weyl group of affine Lie algebra  $\hat\gf$. Using the definition \eqref{eq:8} it is easy to see that $s_{0}s_{i}s_{0}\cdot \lambda=\lambda+\alpha_{i}$, so Weyl group can be presented as semidirect product of Weyl group $W_{\gf}$ of $\gf$ and translations by the roots of $\hat \gf$.

@@

The form \eqref{eq:9} induces the basis dual  to the simple
roots basis. It is called the fundamental weights basis. We denote
its elements by $\omega_i$:
\begin{equation}
  \label{eq:20}
  \langle\omega_i,\alpha_j\rangle=\frac{(\omega_{i},\alpha_{j})}{(\alpha_{j},\alpha_{j})}=\delta_{ij}
\end{equation}
For finite-dimensional Lie algebra there are $r$ fundamental weights, $i=1,\dots, r$. For affine Lie algebra we have additional fundamental weight $\omega_0=\lambda$, $(\lambda,\delta)=1, \; (\lambda,\lambda)=(\delta,\delta)=0$. Other fundamental weights are equal to $\omega_i=a_i^v \lambda_0 +\co{\omega_i}$, where $\co{\omega_i}$ is the fundamental weight of finite-dimensional Lie algebra $\go$.

@@

We consider finitely-generated $\gf$-modules $V$ such that $V=\bigoplus_{\xi\in \hfg^{*}} V_{\xi}$, where each $V_{\xi}$ is finite-dimensional and there exists finite set of weights $\lambda_{1},\dots \lambda_{s}$ which generates weight system of $V$, i.e. if $V_{\xi}\neq 0$ then $\xi=\lambda_{i}-\sum_{k=1,\dots, r} n_{k}\alpha_{k}$ where $n_{k}\in \mathbb{Z}_{+}$ (See \cite{humphreys2008representations}, \cite{carter2005lie}).

@@

The simplest type of highest weight modules is the Verma module
$M^{\mu}$ who's space can be defined as the space
\begin{equation}
  \label{eq:17}
  M^{\mu}=U(\gf)\underset{U(\bff_{+})}{\otimes} D^{\mu}(\bff_{+}),
\end{equation}
with respect to the multiplication in $U(\gf)$,
where $\underset{U(\bff_{+})}{\otimes}$ means that the action of elements of $U(\bff_{+})$ ``falls through'' the left part of tensor product onto the right part. Here $\bff_{+}$ is Borel sub-algebra, $D^{\mu}(\bff_{+})$ is a representation of $\bff_{+}$ such that $D(E^{\alpha})=0,\; D(H)=\mu(H)$ for any positive root $\alpha$.
Elements of $\gf$ act from the left and we should all the elements of $\bff_{+}$ to the right, so that they can act on the space $D_{\lambda}(\bff_{+})$.

Weight multiplicities in Verma module can be found from the Weyl
character formula
\begin{equation}
  \label{eq:11}
  \mathrm{ch} M^{\mu}=\frac{e^{\mu}}{\prod_{\alpha\in \Delta^{+}} \left( 1-e^{-\alpha}\right)^{\mathrm{mult}(\alpha)}}=\frac{e^{\mu}}{\sum_{w\in W} \epsilon(w) e^{w\rho-\rho}}
\end{equation}
Here we have used the Weyl denominator identity
\begin{equation}
  \label{eq:12}
  \prod_{\alpha\in \Delta^{+}} \left( 1-e^{-\alpha}\right)^{\mathrm{mult}(\alpha)}=\sum_{w\in W} \epsilon(w) e^{w\rho-\rho},
\end{equation}
and $\epsilon \left( w\right) :=\det \left( w\right)$ is equal to
the parity of number of Weyl reflections generating $w$.

@@

@@

Each $\gf$-module is also an $\af$-module, although
$L^{\mu}_{\gf}$ is not irreducible as $\af$-module. It can be
decomposed into the sum of irreducible $\af$-modules:
\begin{equation}
  \label{eq:16}
  L^{\mu}_{\gf}=\bigoplus_{\nu}b^{\mu}_{\nu}L^{\nu}_{\af}
\end{equation}
Coefficients in this decomposition are called the branching
coefficients. Using the Weyl character formula we can write a
recurrent relation \cite{2010arXiv1007.0318L} for branching
coefficients $k_{\xi }^{\left( \mu \right) }$ corresponding to an
injection $\af\hookrightarrow \gf$:
\begin{equation}
\begin{array}{c}
k_{\xi }^{\left( \mu \right) }=-\frac{1}{s\left( \gamma _{0}\right) }\left(
\sum_{u\in U}\epsilon (u)\;\dim \left( L_{\afb}^{\mu _{\af%
_{\perp }}\left( u\right) }\right) \delta _{\xi -\gamma _{0},\pi _{%
\widetilde{\af}}(u(\mu +\rho )-\rho )}+\right.  \\
\left. +\sum_{\gamma \in \Gamma _{\widetilde{\af}\rightarrow \gf%
}}s\left( \gamma +\gamma _{0}\right) k_{\xi +\gamma }^{\left( \mu \right)
}\right) .
\end{array}
\label{recurrent-rel}
\end{equation}
The recursion is goverened by the set $\Gamma _{\widetilde{\af}\rightarrow \gf}$ called the injection fan. The latter is defined by the
carrier set $\left\{ \xi \right\} _{\af\rightarrow \gf}$ for the
coefficient function $s(\xi )$
\begin{equation*}
\left\{ \xi \right\} _{\widetilde{\af}\rightarrow \gf}:=\left\{
\xi \in P_{\widetilde{\af}}|s(\xi )\neq 0\right\}
\end{equation*}
appearing in the expansion
\begin{equation}
\prod_{\alpha \in \Delta ^{+}\setminus \Delta _{\bot }^{+}}\left( 1-e^{-\pi
_{\widetilde{\af}}\alpha }\right) ^{\mathrm{mult}(\alpha )-\mathrm{mult}%
_{\af}(\pi _{\widetilde{\af}}\alpha )}=-\sum_{\gamma \in P_{%
\widetilde{\af}}}s(\gamma )e^{-\gamma };\quad
\end{equation}
The weights in $\left\{ \xi \right\} _{\widetilde{\af}\rightarrow \gf}$ are to be shifted by $\gamma _{0}$ -- the lowest vector in $\left\{ \xi
\right\} $ -- and the zero element is to be eliminated:
\begin{equation}
\Gamma _{\widetilde{\af}\rightarrow \gf}=\left\{ \xi -\gamma
_{0}|\xi \in \left\{ \xi \right\} \right\} \setminus \left\{ 0\right\} .
\end{equation}
The formula (\ref{eq:14}) is a particular case of recurrent relation for branching coefficients (\ref{recurrent-rel}) in the case of Cartan subalgebra $\af=\hfg$.

@@

\section{Core datastructures}
\label{sec:core-datastructures} Having introduced necessary
mathematical objects, problems and relations we now describe the
related datastructures introduced in {\bf Affine.m}. Although {\it
Mathematica} is untyped language it is possible to create
structured objects and do type checks with patterns
\cite{shifrinmathematica}, \cite{maeder2000computer}.
\subsection{Weights}
\label{sec:weights}

@@

\begin{lstlisting}[mathescape=true]
  $B_2$ == finiteRootSystem[2, 2, {finiteWeight[2, {1, -1}], finiteWeight[2,{0, 1}]}]}
\end{lstlisting}

@@

Semisimple Lie algebras can be created as sums of simple:
\begin{lstlisting}[mathescape=true]
  $A_1$+$A_1$ == finiteRootSystem[2, 2, {finiteWeight[2, {1, 0}], finiteWeight[2, {0, 1}]}]
\end{lstlisting}

@@

We have implemented several functions to get major properties of root systems. Weyl vector is given by the function \lstinline{rho[rs_?rootSystemQ]}:
\begin{lstlisting}[label=list:1]
  rho[b2]
  finiteWeight[2, {3/2, 1/2}]
\end{lstlisting}
Positive roots can be constructed with \lstinline{positiveRoots[rs_?rootSystemQ]}. For affine Lie algebra this and related functions return the list up to fixed grade. This grade limit is set through the value of \lstinline{rs[gradeLimit]} which is equal to 10 by default. List of roots (up to \lstinline{gradeLimit}) is returned by \lstinline{roots[rs]}. Cartan matrix and fundamental weights are calculated by the functions \lstinline{cartanMatrix} and \lstinline{fundamentalWeights} correspondingly.

@@

Computation of ShortLex form \cite{casselman1994machine,casselman1995automata} for Weyl group elements can be conveniently implemented through the use of pattern-matching features of {\it Mathematica}. In \cite{KallenShortlex} rewrite rules for simple finite dimensional and affine Lie algebras are presented as {\it Mathematica} patterns. Our presentation of Weyl group elements is compatible with the code of \cite{KallenShortlex}
\begin{lstlisting}[mathescape=true]
  In[1]  = $<<$A3reduce;
           reduce[s[1,2,1,2,1,3,2,1,1]]
  Out[1] = s[2, 3, 2]

@@

We represent formal characters of representations by special structure \lstinline{formalElements}. This structure is a hash-table implemented with \lstinline{DownValues}. The keys are weights at the exponents and the values are corresponding multiplicities. \lstinline[mathescape=true]!makeFormalElement[{$\gamma_{1},\dots,\gamma_{n}$},{$m_{1},\dots,m_{n}$}]! creates data-structure which represents the element $\sum_{i=1}^{n} m_{i} e^{\gamma_{i}}$ of formal algebra $\mathcal{E}$. The operations of formal algebra $\mathcal{E}$ are implemented for \lstinline{formalElement} data-type: formal elements can be added, multiplied by number or exponent of weight. There is also multiplication of formal elements but no division.
\begin{lstlisting}[mathescape=true]
  In[1]  = makeFormalElement[{makeFiniteWeight[{1,1}],makeFiniteWeight[{0,0}]},{1,2}] *
             (2 * Exp[makeFiniteWeight[{1,0}]] *
             makeFormalElement[{makeFiniteWeight[{1,1}],makeFiniteWeight[{0,0}]},{1,2}]);
  In[2]  = In[1][weights]
  Out[2] = {finiteWeight[2, {1, 0}], finiteWeight[2, {2, 1}], finiteWeight[2, {3, 2}]}

  In[3]  = In[1][multiplicities]
  Out[3]= {8, 8, 2}
\end{lstlisting}

@@

{\bf Affine.m} can be used to study different kinds of modules, e.g. Verma modules, irreducible modules.  We need datastructure to represent generic module of Lie algebra $\gf$. This datastructure is \lstinline{module}. Module properties can be deduced from its set of singular weights using Weyl character formulae \eqref{eq:11},\eqref{eq:12},\eqref{eq:18},\eqref{eq:13}. Set of singular weights can have Weyl symmetry. It can be symmetry with respect to Weyl group $W_{\gf}$ or with respect to some subalgebra $W_{\af}$ as in the case of parabolic Verma modules. Then it is possible to study only main Weyl chamber $C_{\af}$. So generic constructor for \lstinline{module} datastructure is \lstinline{makeModule[rs_?rootSystemQ][singWeights_formalElement,subs_?rootSystemQ|emptyRootSystem[],opts___?OptionQ]}. Here \lstinline{rs} is root system of Lie algebra $\gf$,\lstinline{singWeights} is the set of singular weights,  \lstinline{subs} is root system corresponding to Weyl group $W_{\af}$ which is the symmetry of the set of singular weights.
There are several constructors for different highest weight modules:
\begin{lstlisting}[mathescape=true]
vm=makeVermaModule[$B_{2}$][weight[$B_{2}$][2,1]];
pm=makeParabolicVermaModule[$B_{2}$][weight[$B_{2}$][2,1],{1}];
im=makeIrreducibleModule[$B_{2}$][weight[$B_{2}$][2,1]];
drawPlaneProjection[2,1,{character[vm],character[pm],character[im]} ]
\end{lstlisting}
\section{Computational algorithms}
\label{sec:comp-algor}

>>>>

\begin{abstract}
%% Text of abstract
In this paper we present {\bf Affine.m} -- program for computations in representation theory of finite-dimensional and affine Lie algebras and describe implemented algorithms.  Algorithms are based upon the properties of weights and Weyl symmetry. The most important for us problems are computation of weight multiplicities in irreducible and Verma modules, branching of representations and tensor product decomposition. These problems have numerous applications in physics and we provide some examples of these applications. The program is implemented in popular computer algebra system {\it Mathematica} and works with finite-dimensional and affine Lie algebras.

@@

{\it Lie algebra} $\gf$ is a vector space with bilinear operation $[\cdot,\cdot]:\gf\otimes\gf\to \gf$, which is called {\it commutator}. If we choose the basis $X_{i}$ in $\gf$ we can specify commutation relations by the {\it structure constants} $C_{ijk}$:
\begin{equation}
  \label{eq:1}
  [X^{i},X^{j}]=\sum_{k} C^{ij}_{k} X^{k}
\end{equation}
Lie algebra is {\it simple} if it contains no non-trivial ideals with respect to commutator. {\it Semisimple} Lie algebra is a direct sum of simple Lie algebras. In present paper we treat simple and semisimple Lie algebras.

Maximal commutative subalgebra ({\it Cartan subalgebra}) of $\gf$ is denoted by $\hfg$.
We denote the elements of basis of $\hfg$ by $H^{i}$.

@@

Although the full set of roots $\Delta$ is infinite for affine Lie algebra $\hat\gf$ the set of simple roots $\Phi$ is finite and its elements are denoted by $\alpha_{0},\dots \alpha_{r}$ where $r=\mathrm{rank}(\gf)$. The roots $\alpha_1,\dots, \alpha_r$ are the roots of the underlying finite-dimensional Lie algebra $\go$. The root $\alpha_0=\delta-\theta$ is the difference of {\it imaginary root} $\delta$ and $\theta$ -- highest root of the algebra $\go$.
Note that root multiplicity $\mathrm{mult}(\alpha)$ for affine Lie algebra can be greater than one.

@@

The Weyl group $W_{\gf}$ is generated by reflections $\{s_{i}:\hfg^{*}\to\hfg^{*}\}$ corresponding to simple roots $\{\alpha_{i}\}$:
\begin{equation}
  \label{eq:8}
  s_{i}\cdot\lambda=\lambda-\frac{2(\alpha_{i},\lambda)}{(\alpha_{i},\alpha_{i})}\alpha_{i}
\end{equation}
Root system and characters of representation are invariant with respect to the action of Weyl group. Root system can be reconstructed from the set of simple roots with the action of Weyl group.

@@

Consider an action of element $s_{0}s_{i}s_{0}$ of Weyl group of affine Lie algebra  $\hat\gf$. Using the definition \eqref{eq:8} it is easy to see that $s_{0}s_{i}s_{0}\cdot \lambda=\lambda+\alpha_{i}$, so Weyl group can be presented as semidirect product of Weyl group $W_{\gf}$ of $\gf$ and translations by the roots of $\hat \gf$. Weyl group element can be presented as the product of elementary reflections in multiple ways. Number of elementary reflections in shortest sequence representing element $w\in W_{\gf}$ is called {\it length} of $w$ and denoted $l(w)$. We also use the notation $\epsilon(w)=(-1)^{l(w)}$ for parity of number of Weyl reflections generating $w$.

@@

The form \eqref{eq:9} induces the basis dual  to the simple
roots basis. It is called the fundamental weights basis. We denote
its elements by $\omega_i$:
\begin{equation}
  \label{eq:20}
  \langle\omega_i,\alpha_j\rangle=\frac{2(\omega_{i},\alpha_{j})}{(\alpha_{j},\alpha_{j})}=\delta_{ij}
\end{equation}
For finite-dimensional Lie algebra there are $r$ fundamental weights, $i=1,\dots, r$. For affine Lie algebra we have additional fundamental weight $\omega_0=\lambda$, $(\lambda,\delta)=1, \; (\lambda,\lambda)=(\delta,\delta)=0$. Other fundamental weights are equal to $\omega_i=a_i^v \lambda_0 +\co{\omega_i}$, where $\co{\omega_i}$ is the fundamental weight of finite-dimensional Lie algebra $\go$.

@@

We consider finitely-generated $\gf$-modules $V$ such that $V=\bigoplus_{\xi\in \hfg^{*}} V_{\xi}$, where each $V_{\xi}$ is finite-dimensional and there exists finite set of weights $\lambda_{1},\dots \lambda_{s}$ which generates weight system of $V$, i.e. if $\mathrm{dim}V_{\xi}\neq 0$ then $\xi=\lambda_{i}-\sum_{k=1,\dots, r} n_{k}\alpha_{k}$ where $n_{k}\in \mathbb{Z}_{+}$ (See \cite{humphreys2008representations}, \cite{carter2005lie}).

@@

The most simple type of highest weight modules is the Verma module
$M^{\mu}$ who's space can be defined as the space
\begin{equation}
  \label{eq:17}
  M^{\mu}=U(\gf)\underset{U(\bff_{+})}{\otimes} D^{\mu}(\bff_{+}),
\end{equation}
with respect to the multiplication in $U(\gf)$,
where $\underset{U(\bff_{+})}{\otimes}$ means that the action of elements of $U(\bff_{+})$ ``falls through'' the left part of tensor product onto the right part. Here $\bff_{+}$ is Borel sub-algebra, $D^{\mu}(\bff_{+})$ is a representation of $\bff_{+}$ such that $D(E^{\alpha})=0,\; D(H)=\mu(H)$ for any positive root $\alpha$.
Elements of $\gf$ act from the left and we should commute all the elements of $\bff_{+}$ to the right, so that they can act on the space $D_{\lambda}(\bff_{+})$.

Weight multiplicities in Verma module can be found from the Weyl
character formula
\begin{equation}
  \label{eq:11}
  \mathrm{ch} M^{\mu}=\frac{e^{\mu}}{\prod_{\alpha\in \Delta^{+}} \left( 1-e^{-\alpha}\right)^{\mathrm{mult}(\alpha)}}=\frac{e^{\mu}}{\sum_{w\in W} \epsilon(w) e^{w\rho-\rho}}
\end{equation}
Here we have used the Weyl denominator identity
\begin{equation}
  \label{eq:12}
  R:=\prod_{\alpha\in \Delta^{+}} \left( 1-e^{-\alpha}\right)^{\mathrm{mult}(\alpha)}=\sum_{w\in W} \epsilon(w) e^{w\rho-\rho},
\end{equation}
and $\epsilon \left( w\right) :=\det \left( w\right)$ is equal to 
the parity of number of Weyl reflections generating $w$.

@@

Each $\gf$-module is also an $\af$-module, although
$L^{\mu}_{\gf}$ is not irreducible as $\af$-module. It can be
decomposed into the sum of irreducible $\af$-modules:
\begin{equation}
  \label{eq:16}
  L^{\mu}_{\gf}=\bigoplus_{\nu}b^{\mu}_{\nu}L^{\nu}_{\af}
\end{equation}
Coefficients in this decomposition are called the branching
coefficients.

Before we proceed to recurrent relation for branching coefficients we need several definitions.

@@

Denote by  $k_{\xi }^{\left( \mu \right) }$ signed branching coefficients. If $\xi\in \bar C_{\af}$ is in main Weyl chamber $k_{\xi}^{(\mu)}=b^{(\mu)}_{\xi}$ otherwise $k_{\xi}^{(\mu)}=\epsilon(w) b^{(\mu)}_{w (\xi+\rho_{\af})-\rho_{\af}}$ where $w\in W_{\af}$ is such that $w (\xi+\rho_{\af})-\rho_{\af}\in W_{\af}$.

Now we can use the Weyl character formula to write a
recurrent relation \cite{2010arXiv1007.0318L} for signed branching
coefficients $k_{\xi }^{\left( \mu \right) }$ corresponding to an
injection $\af\hookrightarrow \gf$:
\begin{equation}
\begin{array}{c}
k_{\xi }^{\left( \mu \right) }=-\frac{1}{s\left( \gamma _{0}\right) }\left(
\sum_{u\in U}\epsilon (u)\;\dim \left( L_{\afb}^{\mu _{\af%
_{\perp }}\left( u\right) }\right) \delta _{\xi -\gamma _{0},\pi _{%
\widetilde{\af}}(u(\mu +\rho )-\rho )}+\right.  \\
\left. +\sum_{\gamma \in \Gamma _{\widetilde{\af}\rightarrow \gf%
}}s\left( \gamma +\gamma _{0}\right) k_{\xi +\gamma }^{\left( \mu \right)
}\right) .
\end{array}
\label{recurrent-rel}
\end{equation}
The recursion is goverened by the set $\Gamma _{\widetilde{\af}\rightarrow \gf}$ called the injection fan. The latter is defined by the
carrier set $\left\{ \xi \right\} _{\af\rightarrow \gf}$ for the
coefficient function $s(\xi )$
\begin{equation*}
\left\{ \xi \right\} _{\widetilde{\af}\rightarrow \gf}:=\left\{
\xi \in P_{\widetilde{\af}}|s(\xi )\neq 0\right\}
\end{equation*}
appearing in the expansion
\begin{equation}
\prod_{\alpha \in \Delta ^{+}\setminus \Delta _{\bot }^{+}}\left( 1-e^{-\pi
_{\widetilde{\af}}\alpha }\right) ^{\mathrm{mult}(\alpha )-\mathrm{mult}%
_{\af}(\pi _{\widetilde{\af}}\alpha )}=-\sum_{\gamma \in P_{%
\widetilde{\af}}}s(\gamma )e^{-\gamma };\quad
\end{equation}
The weights in $\left\{ \xi \right\} _{\widetilde{\af}\rightarrow \gf}$ are to be shifted by $\gamma _{0}$ -- the lowest vector in $\left\{ \xi
\right\} $ -- and the zero element is to be eliminated:
\begin{equation}
\Gamma _{\widetilde{\af}\rightarrow \gf}=\left\{ \xi -\gamma
_{0}|\xi \in \left\{ \xi \right\} \right\} \setminus \left\{ 0\right\} .
\end{equation}
The formula (\ref{eq:14}) is a particular case of recurrent relation for branching coefficients (\ref{recurrent-rel}) in the case of Cartan subalgebra $\af=\hfg$.

@@

\section{Core datastructures}
\label{sec:core-datastructures} Having introduced necessary
mathematical objects, problems and relations we now describe the
related datastructures of {\bf Affine.m}. Although {\it
Mathematica} is untyped language it is possible to create
structured objects and do type checks with patterns
\cite{shifrinmathematica}, \cite{maeder2000computer}.
\subsection{Weights}
\label{sec:weights}

@@

\begin{lstlisting}[mathescape=true]
  $B_2$ == makeFiniteRootSystem[ { {1, -1}, {0, 1} }]
\end{lstlisting}

@@

Semisimple Lie algebras can be created as sums of simple:
\begin{lstlisting}[mathescape=true]
  $A_1\oplus A_1$ == finiteRootSystem[2, 2, {finiteWeight[2, {1, 0}], finiteWeight[2, {0, 1}]}]
\end{lstlisting}

@@

We have implemented several functions to get major properties of root systems. Weyl vector is given by the function \lstinline{rho[rs_?rootSystemQ]}:
\begin{lstlisting}[label=list:1]
  In[1]  =  rho[b2]
  Out[1]  =  finiteWeight[2, {3/2, 1/2}]
\end{lstlisting}
Positive roots can be constructed with \lstinline{positiveRoots[rs_?rootSystemQ]}. For affine Lie algebra this and related functions return the list up to fixed grade. This grade limit is set through the value of \lstinline{rs[gradeLimit]} which is equal to 10 by default. List of roots (up to \lstinline{gradeLimit}) is returned by \lstinline{roots[rs]}. Cartan matrix and fundamental weights are calculated by the functions \lstinline{cartanMatrix} and \lstinline{fundamentalWeights} correspondingly.

@@

Computation of lexicographically minimal form \cite{casselman1994machine,casselman1995automata} for Weyl group elements can be conveniently implemented through the use of pattern-matching features of {\it Mathematica}. In \cite{KallenShortlex} rewrite rules for simple finite dimensional and affine Lie algebras are presented as {\it Mathematica} patterns. Our presentation of Weyl group elements is compatible with the code of \cite{KallenShortlex}
\begin{lstlisting}[mathescape=true]
  In[1]  = $<<$A3reduce;
           reduce[s[1,2,1,2,1,3,2,1,1]]
  Out[1] = s[2, 3, 2]

@@

We represent formal characters of representations by special structure \lstinline{formalElement}. This structure is a hash-table implemented with \lstinline{DownValues}. The keys are weights at the exponents and the values are corresponding multiplicities. \lstinline[mathescape=true]!makeFormalElement[{$\gamma_{1},\dots,\gamma_{n}$},{$m_{1},\dots,m_{n}$}]! creates data-structure which represents the element $\sum_{i=1}^{n} m_{i} e^{\gamma_{i}}$ of formal algebra $\mathcal{E}$. The operations of formal algebra $\mathcal{E}$ are implemented for \lstinline{formalElement} data-type: formal elements can be added, multiplied by number or exponent of weight. There is also multiplication of formal elements but no division.
\begin{lstlisting}[mathescape=true]
  In[1]  = makeFormalElement[{makeFiniteWeight[{1,1}],makeFiniteWeight[{0,0}]},{1,2}] *
             (2 * Exp[makeFiniteWeight[{1,0}]] *
             makeFormalElement[{makeFiniteWeight[{1,1}],makeFiniteWeight[{0,0}]},{1,2}]);
  In[2]  = In[1][weights]
  Out[2] = {finiteWeight[2, {1, 0}], finiteWeight[2, {2, 1}], finiteWeight[2, {3, 2}]}

  In[3]  = In[1][multiplicities]
  Out[3] = {8, 8, 2}
\end{lstlisting}

@@

{\bf Affine.m} can be used to study different kinds of modules, i.e. Verma modules, irreducible modules and parabolic Verma modules .  We need datastructure \lstinline{module} to represent generic module of Lie algebra $\gf$. Module properties can be deduced from its set of singular weights using Weyl character formulae \eqref{eq:11},\eqref{eq:12},\eqref{eq:18},\eqref{eq:13}. Set of singular weights can have Weyl symmetry. It can be symmetry with respect to Weyl group $W_{\gf}$ or with respect to some subalgebra $W_{\af}$ as in the case of parabolic Verma modules. Then it is possible to study only main Weyl chamber $C_{\af}$. To use this symmetry generic constructor for \lstinline{module} datastructure accepts several parameters \lstinline{makeModule[rs_?rootSystemQ][singWeights_formalElement,subs_?rootSystemQ|emptyRootSystem[],opts___?OptionQ]}. Here \lstinline{rs} is root system of Lie algebra $\gf$,\lstinline{singWeights} is the set of singular weights,  \lstinline{subs} is root system corresponding to Weyl group $W_{\af}$ which is the symmetry of the set of singular weights.
There are several specialized constructors for different types of highest weight modules:
\begin{lstlisting}[mathescape=true]
\leabel
vm=makeVermaModule[$B_{2}$][weight[$B_{2}$][2,1]];
pm=makeParabolicVermaModule[$B_{2}$][weight[$B_{2}$][2,1],{1}];
im=makeIrreducibleModule[$B_{2}$][weight[$B_{2}$][2,1]];
Show[textPlot/@{im,vm,pm}]
\end{lstlisting}
\begin{figure}[h]
  \label{irrep-verma-pverma}
  \noindent\centering{
    \includegraphics[width=150mm]{irrep-verma-pverma.pdf}
  }
  \caption{Weight diagrams of $B_{2}$ highest weight modules with highest weight $[2,1]$: irreducible, Verma, parabolic Verma. Produced by code of Listing}
\end{figure}

As we already stated properties of the module are encoded by its singular element. Function \lstinline{singularElement[m_module]} returns singular element of module as \lstinline{formalElement} datastructure. Character (up to \lstinline{limit} for (parabolic) Verma modules) is returned by function \lstinline{character[m_module]}. Direct sum of modules is module and we use natural notation
\begin{lstlisting}[mathescape=true]
im1=makeIrreducibleModule[$B_{2}$][weight[$B_{2}$][2,1]];
im2=makeIrreducibleModule[$B_{2}$][weight[$B_{2}$][1,2]];
character[im1$\oplus$ im2]
\end{lstlisting}
Figure

Tensor product is also implemented but only for finite-dimensional Lie algebras, since tensor product of affine Lie algebra modules leads to rich new structures \cite{kazhdan1994tensor3,kazhdan1993tensor1,kazhdan1993tensor2} which are out of the scope of present paper.

Computation of weight multiplicities in function and branching coefficients 
\section{Computational algorithms}
\label{sec:comp-algor}
